An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-11-04 paper/main.tex
weiter

% Geometry settings
\geometry{
	a4paper,
	total={6in, 9in},
	left=1.25in,
	top=1in,
}

% Title settings
\titleformat{\section}{\normalfont\Large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\normalfont\large\bfseries}{\thesubsection}{1em}{}

% Title
\title{An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network}
\author{Denise Katritschenko, Elias Reich, Sayed Abozar Sadat, Lukas Schwaiger\\
	\small University of Salzburg\\
	\small Department of Artificial Intelligence and Human Interfaces (AIHI) }

@@

\begin{abstract}
	We study how the numerical precision of weights, activations, and gradients in neural
	networks can be reduced without sacrificing too much of the classification
	performance. Using MNIST as a dataset, we benchmark uniform 64-16-bit quantisation,
	mixed-precision formats, and post-training decimal rounding. The results reveal that
	(i) 16-bit end-to-end training does not significantly decrease test accuracy for the
	MNIST dataset and the network used, (ii) three-decimal-place weight rounding while
	keeping 32-bit precision incurs minimal accuracy loss. These findings verify limits
	for memory- and energy-constrained usecases like inference in mobile devices.
\end{abstract}

@@

Artificial Neural Networks (ANNs) are a key part of today’s machine learning
systems. They’ve made huge progress in areas like image recognition, speech

@@

ANNs is that they use a lot of memory and computing power. This becomes a

@@

smartphones, embedded systems, or edge devices. \\ \\ To solve this,

@@

size. But doing this can harm the network’s ability to learn and generalize. A
better strategy is to reduce the numerical precision of the network’s
parameters such as things like weights, activations, and gradients. This
method, called quantization, makes the model smaller and faster while using
less energy, without necessarily hurting accuracy. \\ \\ One advanced version
of this method is called mixed-precision quantization. Instead of using the
same number of bits for every layer in the network, mixed-precision assigns
different bitwidths depending on how important each part is. For example, the
early layers that handle low-level features might use higher precision, while
deeper layers could work fine with fewer bits. This flexible approach often
leads to better results than using a single fixed precision for the whole
model. Researchers have used a range of techniques, like gradient descent,
heuristics, and even evolutionary algorithms, to find the best bitwidth
settings for each layer. \\ \\ Another method that helps reduce resource usage
is network pruning. This means cutting out parts of the network that don’t
contribute much to its performance. One interesting way to do this is with
simulated annealing, a search algorithm that mimics the process of cooling
metal to find a stable structure. It doesn’t need gradient information and can
still find good network shapes by exploring many possibilities. This makes it
useful when you want to shrink a network without going through heavy
retraining. \\ \\ Previous studies have shown that reducing the precision of
neural network parameters can lead to significant improvements in memory usage
and processing speed. Even when precision is reduced to very low levels, models
can still perform competitively on complex tasks. However, while these early
results are promising, the overall effects of lowering precision, especially
how it influences model accuracy, training behavior, and storage efficiency,
are still not fully understood and deserve more detailed investigation. \\ \\
This is where our experimental work begins. In this paper, we take a closer
look at how different levels of parameter precision affect the behavior and
performance of neural networks. Our goal is to understand what happens when we
reduce precision in various ways, and how much we can lower it before
performance starts to drop. \\ \\ We start by testing how reducing precision
during both training and inference affects model accuracy and runtime. Next, we
look at what happens when we train the network using full precision, but lower
the precision only during inference. This setup is common in real-world
applications, where a powerful machine handles training, but the final model is
deployed on a lightweight device. Then, we run a series of tests where we round
the trained weights to a limited number of decimal places. This helps us see
how much we can simplify the stored weight values without hurting accuracy too
much. This is especially useful for devices with very limited storage. We also
try out mixed-precision training, using a mix of FP16 and FP32 formats. On top
of that, we experiment with different types of gradient clipping, like global
norm clipping and value clipping, to see how they affect training stability and
final accuracy. Even though we use relatively simple network architectures and
datasets, our results show that neural networks are quite robust to
low-precision settings. In some cases, we were able to round weights to just
three decimal digits and still get nearly the same accuracy as the
full-precision version. These findings show that it’s possible to significantly
reduce the memory and compute needs of neural networks without losing much
performance. This makes them more practical for use in low-resource settings.
It also opens the door for more advanced techniques that combine quantization,
pruning, and precision-aware training to build highly efficient models for
real-world applications.

\section{Background}

A typical neuron in a neural network performs a weighted sum of its inputs,
followed by an activation function:
\[
	a = \omega^\top x + b,\quad \hat{y} = \sigma(a),
\]
where \( \omega \in \mathbb{R}^n \) is the weight vector, \( b \in \mathbb{R}
\) is the bias, \( x \in \mathbb{R}^n \) is the input, and \( \sigma \) is a
nonlinear activation function (e.g., sigmoid or ReLU).

Training is conducted via backpropagation, where gradients of a loss function
\( L(\hat{y}, y) \) with respect to the model parameters are computed. For
example, using the mean squared error:
% \[
% L(\hat{y}, y) = (\hat{y} - y)^2, \qquad \pdv{L}{\omega_i} = 2(\hat{y} - y) \cdot \odv{\sigma}{a} \cdot x_i.
% \]

\begin{align*}
	L(f(x_1, x_2)) & = {(\sigma(\omega^T x + b) - y)}^2 \\
	               & = {(\sigma(a) - y)}^2              \\
	               & = {(\hat y - y)}^2
\end{align*}
\begin{alignat*}{3}
	\pdv{L}{\omega_1} & = \pdv{L}{\sigma} &  & \pdv{\sigma}{a}      &  & \pdv{a}{\omega_1} \\
	                  & = 2(\hat y - y)~~ &  & \odv{}{a}\sigma(a)~~ &  & x_1
\end{alignat*}

@@

\subsection*{Accuracy Across Bitwidths}
Figure~\ref{fig:modelAccTrainAndInf} shows how the accuracy changed after the
models were trained on the different precision settings. Keep in mind that the
y-axis is not starting from zero, so the absolute height differences are
amplified visually, though the delta is only around $.1\%$. Re-training the
models and re-evaluating them has also lead to the accuracies being
``reversed'', meaning the FP64 model had the least accuracy. It is safe to
assume that the model accuracy has neither decreased, nor increased.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/trainingAndInference.png}
	\caption{Model accuracy evaluated after training with the different precisions.}\label{fig:modelAccTrainAndInf}
\end{figure}

\subsection*{Training and Inference Time}
Also analyzed were the times needed for both the training and inference.
Figure~\ref{fig:modelEvalTrainTimes} depicts the times measured in the training
loop and later on at the inference stage. Again, the times are almost identical
in both regards, and neither a real speed-up nor a slow-down has been measured.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/times.png}
	\caption{Training and evaluation times for the different precisions.}\label{fig:modelEvalTrainTimes}
\end{figure}

@@

\subsection*{Automatic Mixed Precision and Gradient Clipping}
Automatic Mixed Precision (AMP) accelerates deep learning training by using
both FP16 and FP32 data types reducing memory usage and increasing
computational speed. However, FP16’s limited dynamic range can cause numerical
instability, especially during backpropagation. To address this, Gradient
Clipping is used to cap the magnitude of gradients (either globally or per
element) preventing extreme updates that could hinder convergence. Together,
AMP and gradient clipping enable efficient, stable training by balancing
performance and numerical precision.

>>>>

\geometry{
	a4paper,
	total={6in, 9in},
	left=1.25in,
	top=1in,
}

\titleformat{\section}{\normalfont\Large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\normalfont\large\bfseries}{\thesubsection}{1em}{}

\title{An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network}
\author{Denise Katritschenko, Elias Reich, Sayed Abozar Sadat, Lukas Schwaiger\\
	\small University of Salzburg\\
	\small Department of Artificial Intelligence and Human Interfaces (AIHI) }

@@

\begin{abstract}
	\noindent
	We study how the numerical precision of weights, activations, and gradients in
	neural networks can be reduced without sacrificing too much of the performance.
	Using MNIST as a dataset, we benchmark uniform quantisation (mapping real
	values to evenly spaced levels via a single step size), mixed-precision
	formats, and post-training decimal rounding. We compare end-to-end
	low-precision training, inference-only quantisation, and mixed-precision with
	gradient clipping, and evaluate their impact on test accuracy and efficiency.
	The results on the used test setup show that (i) 16-bit end-to-end training
	does not significantly decrease test accuracy for the MNIST dataset, (ii)
	three-decimal-place weight rounding while keeping 32-bit precision incurs
	minimal accuracy loss. These findings verify limits for memory- and
	energy-constrained usecases like inference in mobile devices.
\end{abstract}

@@

Artificial Neural Networks (ANNs) are a key part of today's machine learning
systems. They have made huge progress in areas like image recognition, speech

@@

large ANNs is that they use a lot of memory and computing power. This becomes a

@@

smartphones, embedded systems, or edge devices. \\\\ In order to solve this,

@@

size. But doing this can harm the network's ability to learn and generalize. A
better strategy is to reduce the numerical precision of the network's
parameters such as weights, activations, and gradients. This method, called
quantization, makes the model smaller and faster while using less energy,
without necessarily hurting accuracy. \\\\ An advanced version of this method
is called mixed-precision quantization. Instead of using the same number of
bits for every layer in the network, mixed-precision assigns different
bitwidths depending on how important each part is. For example, the early
layers that handle low-level features might use higher precision, while deeper
layers could work fine with fewer bits. This flexible approach often leads to
better results than using a single fixed precision for the whole model.
Researchers have used a range of techniques, like gradient descent, heuristics,
and even evolutionary algorithms, to find the best bitwidth settings for each
layer. \\\\ Another method that helps reduce resource usage is network pruning.
This means cutting out parts of the network that do not contribute much to its
performance. A way to do this is simulated annealing, a search algorithm that
mimics the process of cooling metal to find a stable structure. It does not
need gradient information and can still find good network shapes by exploring
many possibilities. This makes it useful when you want to shrink a network
without going through heavy retraining. \\\\ Previous studies have shown that
reducing the precision of neural network parameters can lead to significant
improvements in memory usage and processing speed. Even when precision is
reduced to very low levels, models can still perform competitively on complex
tasks. However, while these early results are promising, the overall effects of
lowering precision, especially how it influences model accuracy, training
behavior, and storage efficiency, are still not fully understood and deserve
more detailed investigation. \\\\ This is where our experimental work begins.
In this paper, we take a closer look at how different levels of parameter
precision affect the behavior and performance of neural networks. Our goal is
to understand what happens when we reduce precision in various ways, and how
much we can lower it before performance starts to drop. \\\\ We start by
testing how reducing precision during both training and inference affects model
accuracy and runtime. Next, we look at what happens when we train the network
using full precision, but lower the precision only during inference. This setup
is common in real-world applications, where a powerful machine handles
training, but the final model is deployed on a lightweight device. Then, we run
a series of tests where we round the trained weights to a limited number of
decimal places. This helps to see how much we can simplify the stored weight
values without hurting accuracy too much. This is especially useful for devices
with very limited storage. We also try out mixed-precision training, using a
mix of Half-Precision Floating Point (FP16) and Single-Precision Floating Point
(FP32) formats. On top of that, we experiment with different types of gradient
clipping, like global norm clipping and value clipping, to see how they affect
training stability and final accuracy. \\\\ Even though we use relatively
simple network architectures and datasets, our results show that neural
networks are quite robust to low-precision settings. In some cases, we were
able to round weights to just three decimal digits and still get nearly the
same accuracy as the full-precision version. These findings show that it is
possible to significantly reduce the memory and compute needs of neural
networks without losing much performance. This makes them more practical for
use in low-resource settings. It also opens the door for more advanced
techniques that combine quantization, pruning, and precision-aware training to
build highly efficient models for real-world applications.

\section{Artificial Neural Networks}

A typical neuron in a neural network performs a weighted sum of its inputs,
followed by an activation function (depicted in~\eqref{eq:activation}):

\begin{equation}\label{eq:activation}
	a = \omega^\top x + b,\quad \hat{y} = \sigma(a),
\end{equation}

where $\omega \in \mathbb{R}^n$ is the weight vector, $b \in \mathbb{R}$ is the
bias, $x \in \mathbb{R}^n$ is the input, and $\sigma$ is a nonlinear activation
function (e.g., sigmoid or ReLU). Training is conducted via backpropagation,
where gradients of a loss function $L(\hat{y}, y)$ with respect to the model
parameters are computed. For example, using the mean squared error:

\begin{align}\label{eq:mse}
	L(f(x_1, x_2)) & = {(\sigma(\omega^\top x + b) - y)}^2 \nonumber \\
	               & = {(\sigma(a) - y)}^2 \nonumber                 \\
	               & = {(\hat y - y)}^2
\end{align}
\begin{alignat}{3}\label{eq:grad-w1}
	\pdv{L}{\omega_1} & = \pdv{L}{\sigma} &  & \pdv{\sigma}{a}      &  & \pdv{a}{\omega_1} \nonumber \\
	                  & = 2(\hat y - y)~~ &  & \odv{\sigma(a)}{a}~~ &  & x_1
\end{alignat}

@@

\subsection*{Accuracy Across Bitwidths}
Figure~\ref{fig:modelAccTrainAndInf} shows how the accuracy changed after the
models were trained on the different precision settings. Keep in mind that the
y-axis is not starting from zero, so the absolute height differences are
amplified visually, though the delta is only around $.1\%$. Re-training the
models and re-evaluating them has also lead to the accuracies being
``reversed'', meaning the FP64 model had the least accuracy. It is safe to
assume that the model accuracy has neither decreased, nor increased.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/trainingAndInference.png}
	\caption{Model accuracy evaluated after training with different precisions.}\label{fig:modelAccTrainAndInf}
\end{figure}

\subsection*{Training and Inference Time}
We also analyzed the times needed for both, training and inference.
Figure~\ref{fig:modelEvalTrainTimes} depicts the times measured in the training
loop and later on at the inference stage. Again, the times are almost identical
in both regards, and neither a real speed-up nor a slow-down has been measured.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/times.png}
	\caption{Training and evaluation times for the different precisions.}\label{fig:modelEvalTrainTimes}
\end{figure}

@@

\subsection*{Automatic Mixed Precision and Gradient Clipping}
Automatic Mixed Precision (AMP) accelerates deep learning training by using
both FP16 and FP32 data types reducing memory usage and increasing
computational speed. However, FP16's limited dynamic range can cause numerical
instability, especially during backpropagation. To address this, Gradient
Clipping is used to cap the magnitude of gradients (either globally or per
element) preventing extreme updates that could hinder convergence. Together,
AMP and gradient clipping enable efficient, stable training by balancing
performance and numerical precision.
